---
title: "CAP Bird Community Dynamics - Generalized Linear Mixed Models"
author: "Jeffrey Haight"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r setup, set.seed(54321), include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE, 
  include = TRUE,
  cache = TRUE
  )
rm(list = ls())
gc()
```

```{r packages, message = FALSE, echo = TRUE, warning = FALSE}
  library(tidyverse)
  library(beepr)  # for notifying you when some code is  done running

# for statistical modeling
  library(performance)  # for checking model performance (multicollinearity, VIF, overdispersion, etc.)
  library(insight)      # for extracting a lot of useful model information, using functions like 'get_variance()'
  library(lme4)
  library(glmmTMB)
  library(lmerTest)     # for getting p-values out of the lmer models
  library(MuMIn)        # for calculating R^2, among other things
  library(vegan)        # for multivariate community analyses
  library(jagsUI)       # for running Bayesian models in JAGS
  library(mgcv)         # for GAM(M)s
  library(itsadug)      # for GAMM plotting

# for plotting
# library(ggplot2)      # should already be in the 'tidyverse'
  library(ggpubr)       # for adding correlations and p-values to ggplot
  library(ggeffects)    # for plotting glmm effects in ggplot
  library(ggcorrplot)   # for correlation plots
  library(GGally)       # similar to ggcorrplot
  library(gghighlight)  # for highlighting groups in ggplot
  library(RColorBrewer)
```

# Import Data
```{r, include = FALSE}
list.files("./data")

# This contains data assembled by the '2_1_StatisticalModelingSetup_CAPbirds' file
load("./data/modelinputs_CAPbirds2024.RData")


data.win$site_code <- as.factor(data.spr$site_code)
data.spr$site_code <- as.factor(data.spr$site_code)
```
# COMMUNITY-LEVEL ANALYSIS



### GLMM: Total Bird Abundance

```{r Total Bird Abundance GLMM global model - winter, warning = FALSE}
set.seed(54321)
# cor(data.win$temp_max, data.win$temp_min)
# cor(data.win$temp_max_prevsum, data.win$temp_min_prevsum)

# now try that with different covariates
m.abund.winter <- glmmTMB(abundance ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          # some potential interaction effects
            # year_std * NDBI_std +
            # year_std * LST_std +
            # year_std * LST_prevsum_std +
            # year_std * ppt_sum_std +
            # year_std * ppt_sum_prevsum_std +
            # year_std * NDVI_prevsum_std +
          # (1 |site_code) 
          (1 + year_std|site_code) ##+ (1|id)
          , data = data.win, 
          family = nbinom2
          )
# check_overdispersion(m.abund.winter)
# check_collinearity(m.abund.winter)
# round(confint(m.abund.winter), 3)
# ranef(m.abund.spring)
# r2(m.abund.winter)
summary(m.abund.winter)
AIC(m.abund.winter)
```

```{r Total Bird Abundance GLMM global model - spring, warning = FALSE}
set.seed(54321)
# cor(data.win$temp_max, data.win$temp_min)
# cor(data.win$temp_max_prevsum, data.win$temp_min_prevsum)

# now try that with different covariates
m.abund.spring <- glmmTMB(abundance ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) ##+ (1|id)
          , data = data.spr, 
          family = nbinom2
          )
# check_overdispersion(m.abund.spring)
# check_collinearity(m.abund.spring)
# round(confint(m.abund.spring), 3)
# ranef(m.abund.spring)
# r2(m.abund.spring)
summary(m.abund.spring)
AIC(m.abund.spring)
```
### GLMM: Species Richness
```{r species richness GLMM global model - winter}
set.seed(54321)
m.rich.winter <- glmmTMB(rich ~
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          # some potential interaction effects
            # year_std * NDBI_std +
            # year_std * LST_std +
            # year_std * LST_prevsum_std +
            # year_std * ppt_sum_std +
            # year_std * ppt_sum_prevsum_std +
            # year_std * NDVI_prevsum_std +
          # (1 |site_code) 
          (1 + year_std|site_code) ##+ (1|id)
          , data = data.win, 
          family = poisson    # 'poisson' for richness
          )

# check_overdispersion(m.rich.winter)
# check_collinearity(m.rich.winter)
# round(confint(m.rich.winter), 3)
# ranef(m.rich.winter)
# r2(m.rich.winter)
summary(m.rich.winter)
AIC(m.rich.winter)

```


```{r species richness GLMM global model - spring}
set.seed(54321)
m.rich.spring <- glmmTMB(rich ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) ##+ (1|id)
          , data = data.spr, 
          family = poisson    # 'poisson' for richness
          )

# check_overdispersion(m.rich.spring)
# check_collinearity(m.rich.spring)
# round(confint(m.rich.spring), 3)
# ranef(m.rich.spring)
# r2(m.rich.spring)
summary(m.rich.spring)
AIC(m.rich.spring)
```


# GUILD-LEVEL ANALYSIS
### Variation between seasons
First, we may wonder if there is a seasonal difference in guild abundance. This can also be examined by comparing the intercepts between the two seasonal models (e.g. significant difference if the 95% CIs don't overlap, or if they don't overlap the mean)
```{r seed dispersers - seasonal diff}
wilcox.test(abund_SeedDisp ~ season, data = data.div) #%>% str()
# significant difference between seasons 
```

```{r Insectivores - seasonal diff}
wilcox.test(abund_Invert ~ season, data = data.div)
# no significant difference between seasons
```
```{r pollinators - seasonal diff}
wilcox.test(abund_pollen ~ season, data = data.div)
# significant difference between seasons
```
### Variation among years
```{r variation among years - guilds}
# test for significant variation among years using the Kruskal-Wallis Rank Sum Test
    kruskal.test(abund_SeedDisp ~ survey_year, data = data.win)
    kruskal.test(abund_SeedDisp ~ survey_year, data = data.spr)
    
    kruskal.test(abund_Invert ~ survey_year, data = data.win)
    kruskal.test(abund_Invert ~ survey_year, data = data.spr)
    
    kruskal.test(abund_pollen ~ survey_year, data = data.win)
    kruskal.test(abund_pollen ~ survey_year, data = data.spr)

```


### GLMM: Seed Dispersers

```{r seed dispersers abundance GLMM global model - winter, warning = FALSE}
set.seed(54321)

# now try that with different covariates
m.seed.winter <- glmmTMB(abund_SeedDisp ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) #+  (1|survey_year) #+ (1|id)
          , data = data.win,   # (data.spr %>% filter(abundance < 100))
          family = nbinom2
          )

# check_overdispersion(m.seed.winter)
# check_collinearity(m.seed.winter)
# round(confint(m.seed.winter), 3)
# ranef(m.seed.winter)
# r2(m.seed.winter)
summary(m.seed.winter)
AIC(m.seed.winter)

# plot(ggpredict(m.seed.winter, terms = "year_std"))
```

```{r seed disperser abundance GLMM global model - spring, warning = FALSE}
set.seed(54321)

# now try that with different covariates
m.seed.spring <- glmmTMB(abund_SeedDisp ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) 
          , data = data.spr,
          family = nbinom2
          )

# check_overdispersion(m.seed.spring)
# check_collinearity(m.seed.spring)
# round(confint(m.seed.spring), 3)
# ranef(m.seed.spring)
# r2(m.seed.spring)
summary(m.seed.spring)
AIC(m.seed.spring)
```


### GLMM: Insectivores
```{r invertivore abundance GLMM global model - winter, warning = FALSE}

set.seed(54321)

m.invert.winter <- glmmTMB(abund_Invert ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) 
          , data = data.win,
          family = nbinom2
          )

# check_overdispersion(m.invert.winter)
# check_collinearity(m.invert.winter)
# round(confint(m.invert.winter), 3)
# ranef(m.invert.winter)
# r2(m.invert.winter)
summary(m.invert.winter)
AIC(m.invert.winter)
```
```{r invertivore abundance GLMM global model - spring, warning = FALSE}

set.seed(54321)

# now try that with different covariates
m.invert.spring <- glmmTMB(abund_Invert ~
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          (1 + year_std|site_code) 
          , data = data.spr,
          family = nbinom2
          )

# check_overdispersion(m.invert.spring)
# check_collinearity(m.invert.spring)
# round(confint(m.invert.spring), 3)
# ranef(m.invert.spring)
# r2(m.invert.spring)
summary(m.invert.spring)
AIC(m.invert.spring)
```


### GLMM: Pollinators
Models of pollinator abundance with the random slope effect may fail to converge because there's not enough among-site variance (in trends of pollinator abundance)
```{r pollinator abundance model - winter, warning = FALSE}

set.seed(54321)

m.pol.winter <- glmmTMB(abund_pollen ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          # (1 + year_std|site_code)
          (1 |site_code)
          , data = data.win,
          family = nbinom2
          )

# check_overdispersion(m.pol.winter)
# check_collinearity(m.pol.winter)
# round(confint(m.pol.winter), 3)
# ranef(m.pol.winter)
# r2(m.pol.winter)
# summary(m.pol.winter)
AIC(m.pol.winter)



  
```

```{r pollinator abundance model - spring, warning = FALSE}

set.seed(54321)

m.pol.spring <- glmmTMB(abund_pollen ~ 
          # time
            year_std +   # year as a continuous variable
          # current env. conditions
            NDBI_std +
            NDVI_std +
            LST_std +
            ppt_sum_std +
          # interactions between current env. conditions
            NDBI_std*NDVI_std +
            NDBI_std*LST_std +
            NDBI_std*ppt_sum_std +
          # recent env conditions
            NDVI_prevsum_std +
            LST_prevsum_std +
            ppt_sum_prevsum_std +
          # (1 + year_std|site_code)
          (1 |site_code)
          , data = data.spr,
          family = nbinom2
          )

# check_overdispersion(m.pol.spring)
# check_collinearity(m.pol.spring)
# round(confint(m.pol.spring), 3)
# ranef(m.pol.spring)
# r2(m.pol.spring)
summary(m.pol.spring)
AIC(m.pol.spring)
```


# COMBINE EXPORT MODEL RESULTS



```{r model fixed effects}
rownames(confint(m.pol.winter))

m.coef <- data.frame(
  rbind(
    confint(m.abund.winter),
    confint(m.rich.winter),
    confint(m.seed.winter),
    confint(m.invert.winter),
    confint(m.pol.winter),
    confint(m.abund.spring),
    confint(m.rich.spring),
    confint(m.seed.spring),
    confint(m.invert.spring),
    confint(m.pol.spring)
  )
) %>% rownames_to_column("parameter")

# filter out the random effect parameters
m.coef <- m.coef %>% 
  filter(!grepl('Std.Dev', parameter)) %>% 
  filter(!grepl('Cor.', parameter))  %>% 
  filter(!grepl('NA', parameter)) 


# add the standard errors and p-values to the table
m.coef <- cbind(m.coef, rbind(
    summary(m.abund.winter)$coefficients$cond[,c(2,4)],
    summary(m.rich.winter)$coefficients$cond[,c(2,4)],
    summary(m.seed.winter)$coefficients$cond[,c(2,4)],
    summary(m.invert.winter)$coefficients$cond[,c(2,4)],
    summary(m.pol.winter)$coefficients$cond[,c(2,4)],
    summary(m.abund.spring)$coefficients$cond[,c(2,4)],
    summary(m.rich.spring)$coefficients$cond[,c(2,4)],
    summary(m.seed.spring)$coefficients$cond[,c(2,4)],
    summary(m.invert.spring)$coefficients$cond[,c(2,4)],
    summary(m.pol.spring)$coefficients$cond[,c(2,4)]
  ))


colnames(m.coef) <- c("parameter", "lower95", "upper95", "estimate", "se", "p_value")
  
params <- c("Intercept", "Time (Year)", "Impervious Surface", 
          "Vegetation (Current)", "Temperature (Current)", "Precipitation (Current)",
          "Vegetation (Recent)", "Temperature (Recent)", "Precipitation (Recent)",
          "Impervious Surface X\nVegetation Interaction", 
          "Impervious Surface X\nTemperature Interaction", 
          "Impervious Surface X\nPrecipitation Interaction"#,
          #"Random Intercept by Site (SD)", "Random Time Slope by Site (SD)", "Random Effect Correlation"
          )
# responses <- c("Total Bird Abundance", "Species Richness)
responses <- c("Total Bird Abundance", "Species Richness", 
               "Seed Dispersers", "Insectivores", "Pollinators")
seasons <- c("Winter", "Spring")

n.param <- length(params)
n.responses <- length(responses)
n.seasons <- length(seasons)


m.coef$parameter <- rep(params, n.responses*n.seasons)
m.coef$param_ord <- as.factor(rep(1:n.param, n.responses*n.seasons))
m.coef$param_ord_rev <- as.factor(rep(rev(1:n.param), n.responses*n.seasons))
# m.coef$param_ord <- factor(rep(1:n.param, n.responses*n.seasons),
#                                     levels = rev(params))

m.coef$response <- rep(rep(responses, each = n.param), n.seasons)
m.coef$response_ord <- as.factor(rep(rep(1:n.responses, each = n.param), n.seasons))
m.coef$response_ord_rev <- as.factor(rep(rep(rev(1:n.responses), each = n.param), n.seasons))

m.coef$season <- rep(seasons, each = n.param*n.responses)
m.coef$season <- factor(m.coef$season, levels = c("Winter", "Spring"))

# set the parameter as a factor with levels according to the order you want to plot the fixed effects
m.coef$parameter <- factor(m.coef$parameter, levels = unique(m.coef$parameter))

m.coef

m.coef %>% arrange(lower95)        # lowest range of fixed effects at about -0.63
m.coef %>% arrange(desc(upper95))  # highest range of fixed effects at about -0.48

#
m.coef$CI_95 <- paste(round(m.coef$lower95, 3), round(m.coef$upper95, 3), sep = ", ")
m.coef <- m.coef %>% arrange(response_ord, desc(season))
write.csv(m.coef, "./data/modeloutputs_fixedeffectestimates.csv", row.names = FALSE)
```
```{r model variances, warning = FALSE}
m.r2 <- cbind(
  rep(responses, length(seasons)),
  rep(seasons, each = length(responses)),
  rbind(
    r2(m.abund.winter),
    r2(m.rich.winter),
    r2(m.seed.winter),
    r2(m.invert.winter),
    r2(m.pol.winter),
    r2(m.abund.spring),
    r2(m.rich.spring),
    r2(m.seed.spring),
    r2(m.invert.spring),
    r2(m.pol.spring)
))
m.r2
```

```{r model random effects}
summary(m.abund.winter) %>% str()
```

# PLOT FIXED EFFECTS
```{r abundance effects}
brewer.pal(8, "Dark2")
# Eight colors "#1B9E77" "#D95F02" "#7570B3" "#E7298A" "#66A61E" "#E6AB02" "#A6761D" "#666666"
pal.abund <- c("#E7298A","#66A61E","#E6AB02",  "#666666")

plot.winter <- m.coef %>% filter(!response == "Species Richness") %>% 
  filter(season == "Winter") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate, color = response_ord_rev, fill = response_ord_rev)) +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  geom_linerange(aes(ymin = lower95, ymax = upper95),linewidth = 1, position = position_dodge2(width=0.7)) +
  geom_point(aes(shape = response_ord_rev), size = 1.5, position = position_dodge2(width=0.7), color = "gray20") +
  scale_x_discrete(labels = rev(params[-1])) +
  scale_shape_manual(values = c(23, 22, 24, 21)) +
  # scale_shape_manual(values = c(18, 15, 16, 17)) +
  scale_color_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  scale_fill_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  coord_flip(ylim = c(-0.90, 0.90)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 8),
    axis.title = element_text(face = "bold", size = 10)
  )
plot.winter


plot.spring <- m.coef %>% filter(!response == "Species Richness") %>% 
  filter(season == "Spring") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate, color = response_ord_rev, fill = response_ord_rev)) +
  # theme_minimal() +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  geom_linerange(aes(ymin = lower95, ymax = upper95),linewidth = 1, position = position_dodge2(width=0.7)) +
  geom_point(aes(shape = response_ord_rev), size = 1.5, position = position_dodge2(width=0.7), color = "gray20") +
  scale_x_discrete(labels = rev(params[-1])) +
  scale_shape_manual(values = c(23, 22, 24, 21)) +
  # scale_shape_manual(values = c(18, 15, 16, 17)) +
  scale_color_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  scale_fill_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  coord_flip(ylim = c(-0.9, 0.9)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 8),
    axis.title = element_text(face = "bold", size = 10)
  )
plot.spring

list.files("./figures/effects_env")
ggsave("./figures/effects_env/fig1_modeleffects_abundance_1winter.png",
         plot.winter,
         width = 4,
         height = 4,
         units = "in",
         dpi = 300)

ggsave("./figures/effects_env/fig1_modeleffects_abundance_2spring.png",
       plot.spring,
       width = 5,
       height = 4,
       units = "in",
       dpi = 300)
```

```{r}
plot.both <- m.coef %>% filter(!response == "Species Richness") %>% 
  # filter(season == "2_spring") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate, color = response_ord_rev, fill = response_ord_rev)) +
  # theme_minimal() +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  geom_linerange(aes(ymin = lower95, ymax = upper95),linewidth = 1, position = position_dodge2(width=0.7)) +
  geom_point(aes(shape = response_ord_rev), size = 1, position = position_dodge2(width=0.7), color = "gray20") +
  # geom_pointrange(aes(ymin = lower95, ymax = upper95, shape = response_ord_rev),linewidth = 1, position = position_dodge2(width=0.7), color = "gray20") +
  scale_x_discrete(labels = rev(params[-1])) +  # drop the intercept 
  scale_shape_manual(values = c(23, 22, 24, 21)) +
  # scale_shape_manual(values = c(18, 15, 16, 17)) +
  scale_color_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  scale_fill_manual(labels = rev(responses[c(1,2,3,5)]), values = pal.abund) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  coord_flip(ylim = c(-0.9, 0.9)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # legend.position="bottom", legend.box = "horizontal",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 8),
    axis.title = element_text(face = "bold", size = 10)
  ) + facet_grid(cols = vars(season))
plot.both

ggsave("./figures/effects_env/fig1_modeleffects_abundance_bothseasons.png",
       plot.both,
       width = 6,
       height = 3.75,
       units = "in",
       dpi = 600)
```


```{r}
plot.both <- m.coef %>% filter(response == "Species Richness") %>% 
  # filter(season == "1_winter") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate)) +
  # theme_minimal() +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  geom_linerange(aes(ymin = lower95, ymax = upper95),linewidth = 1, position = position_dodge2(width=0.7), color = "blue3") +
  geom_point(aes(shape = response_ord_rev), size = 2, color = "gray20", fill = "blue3", pch = 21) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  scale_x_discrete(labels = rev(params[-1])) +
  coord_flip(ylim = c(-0.20, 0.20)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # legend.position="bottom", legend.box = "horizontal",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 8),
    axis.title = element_text(face = "bold", size = 10)
  ) + facet_grid(cols = vars(season))
plot.both


ggsave("./figures/effects_env/fig1_modeleffects_richness_bothseasons.png",
       plot.both,
       width = 6,
       height = 3.75,
       units = "in",
       dpi = 600)

```




```{r all response variables - wide}
# figure formatted for the 2024 ESA Annual Meeting poster

plot.both <- m.coef %>%# filter(!response == "Species Richness") %>% 
  # filter(season == "2_spring") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate, color = response_ord_rev, fill = response_ord_rev)) +
  theme_bw() +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.5, linewidth = 0.5, position = position_dodge(width = 0.7), alpha = 0.8) +
  geom_point(aes(shape = response_ord_rev), size = 1.5, position = position_dodge2(width=0.7), color = "gray20") +
  # geom_pointrange(aes(ymin = lower95, ymax = upper95, shape = response_ord_rev),linewidth = 1, position = position_dodge2(width=0.7), color = "gray20") +
  # scale_x_discrete(labels = c("Impervious Surface X Precipitation Interaction", 
  #                             "Impervious Surface X Temperature Interaction", 
  #                             "Impervious Surface X Vegetation Interaction",
  #                             "Precipitation (Recent)" , "Temperature (Recent)", "Vegetation (Recent)", 
  #                             "Precipitation (Current)", "Temperature (Current)","Vegetation (Current)", 
  #                             "Impervious Surface" ,  "Time (Year)")) +  # drop the intercept 
  scale_x_discrete(labels = rev(params[-1])) +  
  scale_shape_manual(values = c(23, 22, 24, 21, 21)) +
  # scale_shape_manual(values = c(18, 15, 16, 17)) +
  scale_color_manual(labels = rev(responses), values = c(pal.abund[c(1,2,3)], "#0000cd", "gray50")) +
  scale_fill_manual(labels = rev(responses), values = c(pal.abund[c(1,2,3)], "#0000cd", "gray50")) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  coord_flip(ylim = c(-0.9, 0.9)) + 
  facet_grid(cols = vars(season)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # legend.position="bottom", legend.box = "horizontal",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 14),
    axis.title = element_text(face = "bold", size = 16)) 
plot.both


ggsave("./figures/effects_env/fig1_modeleffects_allresponses_bothseasons_wide.png",
       plot.both,
       width = 12,
       height = 6,
       units = "in",
       dpi = 600)

```



```{r all response variables - long}
# figure formatted for the 2024 ESA Annual Meeting poster

plot.both <- m.coef %>%# filter(!response == "Species Richness") %>% 
  # filter(season == "2_spring") %>%
  filter(!parameter %in% c("Intercept", "Random Intercept by Site (SD)", 
                           "Random Time Slope by Site (SD)", "Random Effect Correlation")) %>%
ggplot(aes(x = param_ord_rev, y = estimate, color = response_ord_rev, fill = response_ord_rev)) +
  theme_bw() +
  geom_hline(yintercept = 0, linetype="dashed", color = "gray20") +
  # geom_linerange(aes(ymin = lower95, ymax = upper95),linewidth = 0.8, position = position_dodge2(width=0.6)) +
  # geom_errorbar(aes(ymin = lower95, ymax = upper95),linewidth = 0.8, position = position_dodge2(width=0.6)) +
  geom_errorbar(aes(ymin = lower95, ymax = upper95), width = 0.5, linewidth = 0.5, position = position_dodge(width = 0.7), alpha = 0.8) +
  geom_point(aes(shape = response_ord_rev), size = 2, position = position_dodge2(width=0.7), color = "gray20") +
  # geom_pointrange(aes(ymin = lower95, ymax = upper95, shape = response_ord_rev),linewidth = 1, position = position_dodge2(width=0.7), color = "gray20") +
  # scale_x_discrete(labels = c("Impervious Surface X Precipitation Interaction", 
  #                             "Impervious Surface X Temperature Interaction", 
  #                             "Impervious Surface X Vegetation Interaction",
  #                             "Precipitation (Recent)" , "Temperature (Recent)", "Vegetation (Recent)", 
  #                             "Precipitation (Current)", "Temperature (Current)","Vegetation (Current)", 
  #                             "Impervious Surface" ,  "Time (Year)")) +  # drop the intercept 
  scale_x_discrete(labels = rev(params[-1])) +  
  scale_shape_manual(values = c(23, 22, 24, 21, 21)) +
  # scale_shape_manual(values = c(18, 15, 16, 17)) +
  scale_color_manual(labels = rev(responses), values = c(pal.abund[c(1,2,3)], "#0000cd", "gray60")) +
  scale_fill_manual(labels = rev(responses), values = c(pal.abund[c(1,2,3)], "#0000cd", "gray60")) +
  # coord_cartesian(xlim = c(-0.35, 0.35)) +
  coord_flip(ylim = c(-0.9, 0.9)) + 
  facet_grid(cols = vars(season)) +
  labs(x = "Fixed Effect", y = "Estimate")+
  theme(
    legend.position = "none",
    # legend.position="bottom", legend.box = "horizontal",
    # axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8)
    axis.text.y = element_text(size = 8),
    axis.title = element_text(face = "bold", size = 10)) 
plot.both


ggsave("./figures/effects_env/fig1_modeleffects_allresponses_bothseasons_long.png",
       plot.both,
       width = 6.5,
       height = 5.5,
       units = "in",
       dpi = 600)

```


# BONUS
\